\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}
\thispagestyle{myacknowledgements}

This thesis is the result of a long journey filled with constant changes and, ultimately, growth. Starting in late 2020 during a time of social isolation and uncertainty, and concluding in a period of hope, reconnection, and new beginnings, this journey would not have been the same without the support of many people. Some of you I stumbled upon by chance, while others I have known for a long time.

First and foremost, I would like to thank my supervisor, Dr. Guillermo Rodrigo, for his guidance, support, and patience (which I have put to the test on more than one occasion). Since the beginning of my academic career during my Bachelor's thesis in 2018, you have not only been a constant source of inspiration as a scientist but also a mentor and a friend. I would consider myself very lucky if I ever become a fraction of the scientist and person you are.

Secondly, I would like to thank J. Alberto Conejero for his constant support within the university. Although your availability has decreased over the years, I am grateful that you have always found time to help me with my doubts and concerns, even if it meant having phone calls during our commutes home.

I would like to thank my lab mates, with whom I have shared countless hours of work, laughter, and frustration. Special gratitude goes to those of you I have had the pleasure to work with more closely, such as Roser, Raúl, Roswitha, and María. This thesis is as much yours as it is mine.

I would like to thank my Investmat mentors and friends, Antoni and Christian, who adopted this \textit{biotechnologist} as a fellow mathematician and who have given me all the mathematical foundations I needed to survive Investmat, which ultimately led me to this thesis. Perhaps next time there is a conference in Cullera, I can teach you a thing or two about elliptic curves.

Special thanks to Nicolás, who, despite the distance, has always been there to share stories about our misadventures, even when those calls had to happen in the middle of the night. Valencia eagerly awaits your return.

I would also like to thank my Biotech friends, both those who stayed in Valencia - Adrián, Ramón, and Arcadio - for their support and for always being there to hang out every second Thursday or so, even if it meant having to listen to me talk about mathematics, and those who unfortunately did not stay - Jorge and Iñigo. Despite the physical distance between some of us now, I am glad that our bonds remain as strong as they were during our time together from 2015 to 2019 when all six of us shared countless memories.

I would like to thank my friends from L'Alfàs for allowing me to escape from the city and the university every now and then, even if some of you have already left to pursue your dreams. Thank you, Roberto, Adrián, David, Álvaro, and the rest of the gang.

I would also like to thank the Secureum, Spearbit, and webtrES communities for accepting me as one of their own and for enabling me to learn the intricacies of computer science, cybersecurity, and software development. Your friendship and shared experiences have profoundly shaped this thesis. Special thanks to Rajeev from Secureum, who has been a constant source of inspiration, and to my Spanish web3 colleagues Pablo, Eugenio, and Víctor, with whom I spent countless nights hacking, building, and breaking protocols, interspersed with dancing adventures around the world. I am also deeply grateful to the Spearbit/Cantina team, especially Hari, Alex, and Noah, who not only gave me the opportunity to contribute to the core team but also, unknowingly, mentored me in various soft skills.

Finally, to my parents, who were calling me ``Dr.'' right after I started my PhD journey. Maybe academia wasn't that bad after all, was it?